---
title: "Microbiome Data Simulation"
author: "Dahun Seo"
date: "2024-12-11"
geometry: "top=2cm, bottom=2cm, left=2.5cm, right=2.5cm"
header-includes:
# - \usepackage[hangul]{kotex}
# - \setmainhangulfont{나눔고딕}
- \setmainfont{Microsoft Sans Serif}
output:
  pdf_document: 
    latex_engine: xelatex
    highlight: zenburn
    toc: true
    toc_depth: 2
    number_sections: true
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = T, comment = "#>")
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 50), tidy = TRUE)
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

```{r message=FALSE, results="hide", include=FALSE}
rm(list = ls())
```

```{r message=FALSE, include=FALSE}
suppressPackageStartupMessages({
    library(MASS)
    library(dirmult)
    library(ape)
})

source("/Users/dahunseo/programming/microbiome-logcontrast/Simulation/functions.R")
source("/Users/dahunseo/programming/microbiome-logcontrast/Biometrika_2014/ConstLasso.R")
```

# Data Generation

## The logistic normal (LN) distribution

- generate random binary tree with $p$ variables
- calculate cophenetic distanace between variables
- define variance-covariance matrix using distance matrix
\[
    \Sigma_{ij} = \exp(-d_{ij}) / 2
\]
- generate data from multivariate normal distribution
\[
    M_i \sim \mathcal{N}_p(\alpha_0, \Sigma)
\]
- transformation
\[
    Z_{ij} = \log \left(  \frac{\exp(M_{ij})}{\sum_{j=1}^p \exp(M_{ij})} \right)
\]
- generate outcome variable
\[
    Y_i = \beta_0 + Z_i^\intercal \beta + \varepsilon
\]
where $\varepsilon \sim \mathcal{N}(0, \sigma^2)$.

```{r}
set.seed(1234)

# The logistic normal (LN) distribution
n <- 50            # n: sample size
p <- 25             # p: number of features
noise_sigma <- 1    # noise_sigma: noise level for response

# parameters for normal distribution
# base_mu <- rep(0, p)

# parameters for normal distribution
# Some taxa are often significantly more abundant than others, as commonly seen in real microbiome compositional data.
base_mu <- c(rep(p/2, 5), rep(1, p-5)) 

# Create a random binary tree for the p features
random_tree <- ape::rcoal(p)
```

```{r fig.align="center", fig.width=8, fig.height=8}
tree_info <- create_tree_structure(tips = 1:p, edges = random_tree$edge)
tree_info$levels

g <- igraph::graph_from_edgelist(random_tree$edge, directed = FALSE)

layout <- igraph::layout_as_tree(g, root = tree_info$levels[[length(tree_info$levels)]], mode = "out")

plot(g,
    layout = layout,
    vertex.label = igraph::V(g)$name,  
    vertex.size = 5,
    vertex.label.cex = 0.8,         
    vertex.color = "skyblue", 
    edge.color = "black"
)
```

```{r}
# Compute the distance matrix using cophenetic distances
dist_matrix <- stats::cophenetic(random_tree)

# variance-covariance setting
sigma <- exp(-dist_matrix) / 2

# calculate mu for each i
mu_matrix <- matrix(NA, n, p)
for (i in 1:n) {
    mu_matrix[i, ] <- base_mu
}

# generation of logistic normal samples
z <- matrix(NA, n, p)
for (i in 1:n) {
    normal_sample <- MASS::mvrnorm(1, mu_matrix[i, ], sigma)
    exp_sample <- exp(normal_sample)
    z[i, ] <- exp_sample / sum(exp_sample)
}

# label
colnames(z) <- paste0("x", 1:p)

# add the psudeo-count 0.5
z <- ifelse(z == 0, 0.5, z)
```

```{r}
# check
head(z)
apply(z, 1, sum)
```

```{r}
# log transformation
log_z <- log(z)

# coefficients
beta_non_zero <-
    c(
        -3, 3, 2.5, -1, -1.5,
        3, 3, -2, -2, -2,
        1, -1, 3, -2, -1
    )

if (length(beta_non_zero) >= p) {
    beta <- beta_non_zero[1:p]
} else {
    beta <- c(beta_non_zero, rep(0, p - length(beta_non_zero)))
}

base_y <- rep(100, n)

y <- base_y + as.vector(log_z %*% beta) + stats::rnorm(n, 0, sd = noise_sigma)
```

\clearpage

# Simulation Study Results

## Original

```{r fig.align="center", out.width="60%"}
expanded_z <- construct_features(z, tips = 1:p, edges = random_tree$edge, eta = 0, tau = 1)

C <- matrix(1, p, ncol = 1)

res2 <- ConstrLassoCrossVal(y = y, x = log_z, C = C, nfolds = 10)

plot(res2$cvm)

res2$bet.sel
t(C) %*% res2$bet.sel

res2$cvm[res2$sel]
res2$Rsq.sel

{plot(
    beta, res2$bet.sel,
    xlab = "true beta", 
    ylab = "estimated beta",
    pch = 16,
    col = "blue",
    asp = 1
)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)}

{plot(
    y, log_z %*% res2$bet.sel + res2$int.sel,
    xlab = "true y", 
    ylab = "y hat",
    pch = 16,
    col = "blue",
    asp = 1
)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)}
```

\clearpage

## Expanded

```{r fig.align="center", out.width="60%"}
head(expanded_z$data)

res3 <- ConstrLassoCrossVal(y = y, x = expanded_z$data, C = expanded_z$C, nfolds = 10)

plot(res3$cvm)

res3$bet[, which.min(res3$cvm)]

t(expanded_z$C) %*% res3$bet[, which.min(res3$cvm)]

res3$cvm[res3$sel]
res3$Rsq.sel

{plot(
    y, expanded_z$data %*% res3$bet.sel + res3$int.sel,
    xlab = "true y", 
    ylab = "y hat",
    pch = 16,
    col = "blue",
    asp = 1
)
abline(a = 0, b = 1, col = "red", lty = 2, lwd = 2)}
```